### 2016-03-14 paper/icfp16-submission/semantics.tex
update proofs
comment out unused lemmas

%   % \force{v}{\subst{\tsu}{\thole}}{\vsu}{\tsu} & \mbox{if}\ \thole \in dom(\tsu) \\
%   % \triple{v}{\vsu}{\extendsu{\tsu}{\thole}{\typeof{v}}} & \mbox{otherwise} \\
% \end{cases} \\
\force{\vhole{\thole}}{t}{\vsu}{\tsu} & \defeq & \hspace{-1ex}
\begin{cases}
  \triple{v}{\vsu}{\tsu'}    & \mbox{if } v = \lookupsu{\vsu}{\ehole},
                                         \tsu' = \unify{\{\thole, t, \typeof{v}\}}{\tsu}\\
  \triple{\stuck}{\vsu}{\tsu} & \mbox{if } v = \lookupsu{\vsu}{\ehole} \\
  \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu'} & \mbox{if}\ \tsu' = \unify{\{\thole, t\}}{\tsu}, v = \gen{t}{\tsu'}\\
\end{cases} \\
% \begin{cases}
%   \triple{\lookupsu{\vsu}{\ehole}}{\vsu}{\tsu} & \mbox{if}\ \ehole \in dom(\vsu), \hastype{\lookupsu{\vsu}{\ehole}}{t} \\
%   \triple{\stuck}{\vsu}{\tsu}                  & \mbox{if}\ \ehole \in dom(\vsu) \\
%   \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu}   & v = \gen{t}\\
% \end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \triple{\vsu(i)}{\vsu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\triple{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{n}{\tint}{\vsu}{\tsu}     & \defeq & \triple{n}{\vsu}{\tsu} \\
\force{b}{\tbool}{\vsu}{\tsu}    & \defeq & \triple{b}{\vsu}{\tsu} \\
\force{\efun{x}{e}}{\tfun}{\vsu}{\tsu} & \defeq & \triple{\efun{x}{e}}{\vsu}{\tsu} \\
\force{\epair{v_1}{v_2}}{\tprod{t_1}{t_2}}{\vsu}{\tsu} & \defeq & \triple{\epair{v_1}{v_2}}{\vsu}{\tsu''}, \mbox{if}\ \tsu' = \unify{\{\typeof{v_1}, t_1\}}{\tsu}, \tsu'' = \unify{\{\typeof{v_2}, t_2\}}{\tsu'}  \\
\force{\vleaf{t_1}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vleaf{t_1}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu} \\
\force{\vnode{t_1}{v_1}{v_2}{v_3}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vnode{t_1}{v_1}{v_2}{v_3}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu} \\
\force{v}{t}{\vsu}{\tsu} & \defeq & \triple{\stuck}{\vsu}{\tsu}
\end{array}
$$
\caption{Narrowing values}
\label{fig:narrow}
\end{figure*}
%
%While a hole may map to a value that \emph{contains} another hole, \eg a
%lambda or a tree, it may not map \emph{directly} to another hole,
As the value substitution is normalized, in the first case of \forcesym\ we
do not need to \forcesym\ the result of the substitution, the sub-hole
will be narrowed when the context demands it.

@@

A key technical challenge in generating witnesses is
that we have no (static) type information to rely upon.
%
Thus, we must avoid the trap of generating \emph{spurious}
witnesses that arise from picking irrelevant values, when
instead there exist perfectly good values of a \emph{different}
type, under which the program would not have gone wrong.
%
We now show that our evaluation relation instantiates holes
in a \emph{general} manner. That is, given a lambda-term $f$,
if we have $\steps{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,
then \emph{for every} input type $t$, we can find a value
$v$ of type $t$ such that $\eapp{f}{v}$ goes wrong.

@@

\paragraph{Dynamic Type Inference}
We can think of the evaluation of \eapp{f}{\vhole{\thole}}
as synthesizing a partial instantiation of \thole, and hence,
\emph{dynamically inferring} a (partial) type for $f$'s input.
%
We can extract this type from an evaluation trace, by
\emph{resolving} the \thole\ with the final type
substitution at the end of the trace.
%
Formally, we say that if
$\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\estuck}{\vsu}{\tsu}$,
then the \emph{partial input type} of $f$ upto $\trace$, written
\ptype{\trace}{f}, is $\resolve{\thole}{\tsu}$.

@@

\paragraph{Narrowing}
%
Only a successful call to \forcesym can change the partial
input type of $f$.
%
\begin{lem}
\label{lem:force-inst}
If
$\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$
and
$\trace' \defeq \trace, \step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$
(\ie $\trace'$ is a single-step extension of $\trace$)
and
$\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
then the final step $\step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$
 successfully invokes \forcesym.
%
\ES{not clear we still need this}
\end{lem}

\begin{proof}
  By case analysis on the evaluation rules.
  %
  If $\ptype{\trace}{f} \neq \ptype{\trace'}{f}$ then,
  % one of the holes in $f$'s
  % argument must have been instantiated with a concrete value at the last step.
  by the definition of $\ptype{\trace}{f}$, $\tsu \neq \tsu'$, as \thole
  does not change.
  %
  % An examination of the rules shows that only place this happens is
  % in the second case of \forcesym.
  An examination of the rules shows that only \forcesym can update \tsu,
  and furthermore that only the successful cases of \forcesym do update
  \tsu.
\end{proof}

@@

\paragraph{Preservation}
We prove that each evaluation step \emph{refines} the partial input type
of $f$, \ie\ preserves type compatibility.
%
\begin{lem}
\label{lem:refine-partial}
If $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$ and
$\trace'$ is a single-step extension of $\trace$, % \defeq \trace, \triple{\estuck'}{\vsu'}{\tsu'}$
%
%The partial type of $f$ upto $\trace$ is compatible
%with the partial type upto $\trace'$, \ie\
%
then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
%
\ES{not clear we still need this}
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules.
  %
  First note that by Lemma~\ref{lem:force-inst} we can immediately
  discharge the \rulename{E-*-Bad} rules as they cannot change
  \ptype{\trace}{f} at all, and are thus trivially
  compatibility-preserving. For the \rulename{E-*-Good} rules we can
  show that, by virtue of \forcesym succeeding, all must preseve
  compatibility.
  % Note that all rules preserve partial types with the exception of when
  % \forcesym\ is called on a hole, in which case we may instantiate the hole with
  % a concrete value.
  % %
  % But $\typeof{\vhole{\thole}} = \thole$, which is compatible with any type.
\end{proof}

@@

We prove that the partial instantiation of a type hole $\thole$
is always a refinement of the partial instantiation of the
associated value hole $\ehole$.

\begin{lem}
\label{lem:resolve-compat}
For all traces
$\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$,
$\vsub{\resolve{\thole}{\tsu}}{\resolve{\ehole}{\vsu}}$.
\end{lem}
\begin{proof}
  By induction on $\trace$.
  %
  In the base case $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}$
  and $\thole$ is trivially a refinement of $\vhole{\thole}$.
  %
  In the inductive case, consider the single-step extension of $\trace$,
  $\trace' = \trace,\triple{e'}{\vsu'}{\tsu'}$.
  %
  We show by case analysis on the evaluation rules that if
  $\vsub{\resolve{\thole}{\tsu}}{\resolve{\ehole}{\vsu}}$, then
  $\vsub{\resolve{\thole}{\tsu'}}{\resolve{\ehole}{\vsu'}}$.
  %
\end{proof}

\begin{lem}
\label{lem:vsu-ext}
If  $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$
and $\trace'$ is a single-step extension of $\trace$
and $\ptype{\trace}{f} \neq \ptype{\trace'}{f}$
% and $\resolve{\thole}{\tsu} \neq \resolve{\thole}{\tsu'}$
then $\tsu' = \tsu[\thole_1 \mapsto t_1] \ldots [\thole_n \mapsto t_n]$.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules.
  %
  $\thole$ does not change, so if the partial input types differ then
  $\tsu \neq \tsu'$.
  %
  Note that only \forcesym can change \tsu, and it can only do so
  via \unifysym, which can only extend \tsu.
  % and only the first case,
  % when we extend \vsu\ with $[\ehole \mapsto v]$,
  % does.
  %
  \ES{fill in cases in the appendix}
\end{proof}

Next, we introduce a notion of type % and value
contexts, defined analogously to evaluation contexts.
%
$$
\begin{array}{lcl}
  T &::=& \bullet \spmid \ttree{T} \\
  % V &::=& \bullet \spmid \vnode{t}{V}{v}{v} \spmid \vnode{t}{v}{V}{v} \spmid \vnode{t}{v}{v}{V} \\
\end{array}
$$
%

@@

  % By Lemma~\ref{lem:fixme} we know that
  % $\vcompat{\resolve{\vhole{\thole}}{\vsu}}{\resolve{\thole}{\tsu}}$.
  %
  By Lemma~\ref{lem:vsu-ext} we know that
  $\tsu_{i} = \tsu_{i-1}[\thole_1 \mapsto t_1] \ldots [\thole_n \mapsto t_n]$.
  %
  We will assume, without loss of generality, that
  $\tsu_{i} = \tsu_{i-1}[\thole' \mapsto t']$.
  %
  Since $\tsu_{i-1}$ and $\tsu_{i}$ differ only in $\thole'$ but the resolved
  types differ,
  $\ptype{\trace_{i-1}}{f} = \intctx{\thole'}$
  and
  $\ptype{\trace_i}{f} = \intctx{t'}$.
  % $\resolve{\thole}{\tsu_{i-1}} = \intctx{\thole'}$
  % and
  % $\resolve{\thole}{\tsu_{i}} = \intctx{t'}$.
  %
  By Lemma~\ref{lem:context-compat},
  let $s$ be a concrete type such that $\intctx{s} = t$.
  %
  We show by case analysis on the evaluation rules that
  %
  $$\step{e_{i-1}}{\vsu_{i-1}}{\tsu_{i-1}[\thole' \mapsto s]}{\stuck}{\vsu}{\tsu}$$
  %
  Finally, by Lemma~\ref{lem:resolve-compat} we know that
  % $\vsub{\resolve{\thole}{\tsu_{i-1}}}{\resolve{\ehole}{\vsu_{i-1}}}$,
  $\vsub{\ptype{\trace_{i-1}}{f}}{\resolve{\ehole}{\vsu_{i-1}}}$,
  thus $\vsub{\intctx{\thole'}}{\resolve{\ehole}{\vsu_{i-1}}}$.
  %
  Let $u = \gen{s}{\tsu}$
  and $v = \resolve{\ehole}{\vsu_{i-1}}\sub{\ehole'[\thole']}{u}\sub{\thole'}{s}$,
  $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$ in $i$ steps.
  %and
  % $\vsub{\resolve{\thole}{\tsu_{i}}}{\resolve{\ehole}{\vsu_{i}}}$.
  % \vcompat{\resolve{\vhole{\thole}}{\vsu_{i-1}}}{\resolve{\thole}{\tsu_{i-1}}}
  % =
  % \tcompat{\ptype{i-1}{f}}{t}
  % \\
  % \vcompat{\resolve{\vhole{\thole}}{\vsu_{i}}}{\resolve{\thole}{\tsu_{i}}}
  % =
  % \tincompat{\ptype{i}{f}}{t}
  %
  % Finally, by the definition of $\typeofsym$,
  % we know that $\resolve{\vhole{\thole}}{\vsu_{i-1}} = \invctx{v}$
  % such that $\vsub{T'[\thole']}{v}$.
  %such that $\resolve{\typeof{v}}{\tsu_{i-1}} = T'[\thole']$
  %
  %\ES{need to connect \ptype{\trace{i}}{f} and \resolve{\vhole{\thole}}{\vsu_i}}
  %
  % Now, consider any value $u$ such that $\vcompat{\invctx{u}}{t}$,
  % we show by case analysis on the evaluation relation that
  % $\invctx{u}$ could not make progress at this step, and thus
  % $\steps{\eapp{f}{\invctx{u}}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
\end{proof}

@@

\begin{proof}[\textbf{Proof of Theorem~\ref{thm:soundness}}]
%
Suppose $\trace$ witnesses that $f$ gets stuck,
and let $s = \ptype{\trace}{f}$.
We show that \emph{all} types $t$ have stuck-inducing
values by splitting cases on whether $t$ is
compatible with $s$. %the partial type upto $\trace$.
%
\begin{description}
\item [Case \tcompat{s}{t}:]
  Let $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{\stuck}{\vsu}{\tsu}$.
  %
  The value $v = \resolve{\ehole}{\vsu}$ demonstrates that
  $\eapp{f}{v}$ gets stuck.
\item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, we can derive
  a $v$ from $\trace$ such that $\hastype{v}{t}$ and $\eapp{f}{v}$ gets
  stuck.
% \item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, every $v$
%   such that \hastype{v}{t} demonstrates that $\eapp{f}{v}$ gets stuck.
%   % \ES{do we need to say anythign else?}
\end{description}
\end{proof}

>>>>

%   % \force{v}{\subst{\tsu}{\thole}}{\vsu}{\tsu} & \mbox{if}\ \thole \in dom(\tsu) \\
%   % \triple{v}{\vsu}{\extendsu{\tsu}{\thole}{\typeof{v}}} & \mbox{otherwise} \\
% \end{cases} \\
\force{\vhole{\thole}}{t}{\vsu}{\tsu} & \defeq & \hspace{-1ex}
\begin{cases}
  \triple{v}{\vsu}{\tsu'}    & \mbox{if } v = \lookupsu{\vsu}{\vhole{\thole}},
                                         \tsu' = \unify{\{\thole, t, \typeof{v}\}}{\tsu}\\
  \triple{\stuck}{\vsu}{\tsu} & \mbox{if } v = \lookupsu{\vsu}{\vhole{\thole}} \\
  \triple{v}{\extendsu{\vsu}{\vhole{\thole}}{v}}{\tsu'} & \mbox{if}\ \tsu' = \unify{\{\thole, t\}}{\tsu}, v = \gen{t}{\tsu'}\\
\end{cases} \\
% \begin{cases}
%   \triple{\lookupsu{\vsu}{\ehole}}{\vsu}{\tsu} & \mbox{if}\ \ehole \in dom(\vsu), \hastype{\lookupsu{\vsu}{\ehole}}{t} \\
%   \triple{\stuck}{\vsu}{\tsu}                  & \mbox{if}\ \ehole \in dom(\vsu) \\
%   \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu}   & v = \gen{t}\\
% \end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \triple{\vsu(i)}{\vsu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\triple{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{n}{\tint}{\vsu}{\tsu}     & \defeq & \triple{n}{\vsu}{\tsu} \\
\force{b}{\tbool}{\vsu}{\tsu}    & \defeq & \triple{b}{\vsu}{\tsu} \\
\force{\efun{x}{e}}{\tfun}{\vsu}{\tsu} & \defeq & \triple{\efun{x}{e}}{\vsu}{\tsu} \\
\force{\epair{v_1}{v_2}}{\tprod{t_1}{t_2}}{\vsu}{\tsu} & \defeq & \triple{\epair{v_1}{v_2}}{\vsu}{\tsu''}, \mbox{if}\ \tsu' = \unify{\{\typeof{v_1}, t_1\}}{\tsu}, \tsu'' = \unify{\{\typeof{v_2}, t_2\}}{\tsu'}  \\
\force{\vleaf{t_1}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vleaf{t_1}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu} \\
\force{\vnode{t_1}{v_1}{v_2}{v_3}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vnode{t_1}{v_1}{v_2}{v_3}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu} \\
\force{v}{t}{\vsu}{\tsu} & \defeq & \triple{\stuck}{\vsu}{\tsu}
\end{array}
$$
\caption{Narrowing values}
\label{fig:narrow}
\end{figure*}
%
%While a hole may map to a value that \emph{contains} another hole, \eg a
%lambda or a tree, it may not map \emph{directly} to another hole,
As the value substitution is normalized, in the first case of \forcesym\ we
do not need to \forcesym\ the result of the substitution, the sub-hole
will be narrowed when the context demands it.

@@

A key technical challenge in generating witnesses is
that we have no (static) type information to rely upon.
%
Thus, we must avoid the trap of generating \emph{spurious}
witnesses that arise from picking irrelevant values, when
instead there exist perfectly good values of a \emph{different}
type, under which the program would not have gone wrong.
%
We now show that our evaluation relation instantiates holes
in a \emph{general} manner. That is, given a lambda-term $f$,
if we have $\steps{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,
then \emph{for every} concrete type $t$, we can find a value
$v$ of type $t$ such that $\eapp{f}{v}$ goes wrong.

@@

\paragraph{Dynamic Type Inference}
We can think of the evaluation of \eapp{f}{\vhole{\thole}}
as synthesizing a partial instantiation of \thole, and hence,
\emph{dynamically inferring} a (partial) type for $f$'s input.
%
We can extract this type from an evaluation trace by
applying the final type substitution to \thole.
% \emph{resolving} the \thole\ with the final type
% substitution at the end of the trace.
%
Formally, we say that if
$\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\estuck}{\vsu}{\tsu}$,
then the \emph{partial input type} of $f$ up to $\trace$, written
\ptype{\trace}{f}, is $\resolve{\thole}{\tsu}$.

@@

% \paragraph{Narrowing}
% %
% Only a successful call to \forcesym can change the partial
% input type of $f$.
% %
% \begin{lem}
% \label{lem:force-inst}
% If
% $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$
% and
% $\trace' \defeq \trace, \step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$
% (\ie $\trace'$ is a single-step extension of $\trace$)
% and
% $\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
% then the final step $\step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$
%  successfully invokes \forcesym.
% %
% \ES{not clear we still need this}
% \end{lem}

% \begin{proof}
%   By case analysis on the evaluation rules.
%   %
%   If $\ptype{\trace}{f} \neq \ptype{\trace'}{f}$ then,
%   % one of the holes in $f$'s
%   % argument must have been instantiated with a concrete value at the last step.
%   by the definition of $\ptype{\trace}{f}$, $\tsu \neq \tsu'$, as \thole
%   does not change.
%   %
%   % An examination of the rules shows that only place this happens is
%   % in the second case of \forcesym.
%   An examination of the rules shows that only \forcesym can update \tsu,
%   and furthermore that only the successful cases of \forcesym do update
%   \tsu.
% \end{proof}

@@

% \paragraph{Preservation}
% We prove that each evaluation step \emph{refines} the partial input type
% of $f$, \ie\ preserves type compatibility.
% %
% \begin{lem}
% \label{lem:refine-partial}
% If $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$ and
% $\trace'$ is a single-step extension of $\trace$, % \defeq \trace, \triple{\estuck'}{\vsu'}{\tsu'}$
% %
% %The partial type of $f$ upto $\trace$ is compatible
% %with the partial type upto $\trace'$, \ie\
% %
% then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
% %
% \ES{not clear we still need this}
% \end{lem}
% \begin{proof}
%   By case analysis on the evaluation rules.
%   %
%   First note that by Lemma~\ref{lem:force-inst} we can immediately
%   discharge the \rulename{E-*-Bad} rules as they cannot change
%   \ptype{\trace}{f} at all, and are thus trivially
%   compatibility-preserving. For the \rulename{E-*-Good} rules we can
%   show that, by virtue of \forcesym succeeding, all must preseve
%   compatibility.
%   % Note that all rules preserve partial types with the exception of when
%   % \forcesym\ is called on a hole, in which case we may instantiate the hole with
%   % a concrete value.
%   % %
%   % But $\typeof{\vhole{\thole}} = \thole$, which is compatible with any type.
% \end{proof}

@@

We prove that the partial instantiation of a type hole $\thole$
is always a refinement of the partial instantiation of the
associated value hole $\vhole{\thole}$.

\paragraph{Preservation}
%
We prove two preservation lemmas. First, we show that each evaluation
step refines the partial input type of $f$, thus preserving type
compatibility.
%
\begin{lem}
\label{lem:vsu-ext}
If  $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$
and $\trace'$ is a single-step extension of $\trace$
and $\ptype{\trace}{f} \neq \ptype{\trace'}{f}$
% and $\resolve{\thole}{\tsu} \neq \resolve{\thole}{\tsu'}$
then $\tsu' = \tsu[\thole_1 \mapsto t_1] \ldots [\thole_n \mapsto t_n]$.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules.
  %
  $\thole$ does not change, so if the partial input types differ then
  $\tsu \neq \tsu'$.
  %
  Note that only \forcesym can change \tsu, and it can only do so
  via \unifysym, which can only extend \tsu.
  % and only the first case,
  % when we extend \vsu\ with $[\ehole \mapsto v]$,
  % does.
  %
  % \ES{fill in cases in the appendix}
\end{proof}
%
Second, we show that at each step of evaluation, the partial input type of $f$
is a refinement of the instantiation of $\vhole{\thole}$.
%
\begin{lem}
\label{lem:resolve-compat}
For all traces
$\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$,
$\vsub{\ptype{\trace}{f}}{\resolve{\vhole{\thole}}{\vsu}}$.
\end{lem}
\begin{proof}
  By induction on $\trace$.
  %
  In the base case $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}$
  and $\thole$ is trivially a refinement of $\vhole{\thole}$.
  %
  In the inductive case, consider the single-step extension of $\trace$,
  $\trace' = \trace,\triple{e'}{\vsu'}{\tsu'}$.
  %
  We show by case analysis on the evaluation rules that if
  $\vsub{\ptype{\trace}{f}}{\resolve{\vhole{\thole}}{\vsu}}$, then
  $\vsub{\ptype{\trace'}{f}}{\resolve{\vhole{\thole}}{\vsu'}}$.
  %
\end{proof}

\ES{Where does this fit in??}
Next, we introduce a notion of type % and value
contexts, defined analogously to evaluation contexts.
%
$$
\begin{array}{lcl}
  T &::=& \bullet \spmid \ttree{T} \spmid \tprod{T}{t} \spmid \tprod{t}{T} \\
  % V &::=& \bullet \spmid \vnode{t}{V}{v}{v} \spmid \vnode{t}{v}{V}{v} \spmid \vnode{t}{v}{v}{V} \\
\end{array}
$$
%

@@

  % By Lemma~\ref{lem:fixme} we know that
  % $\vcompat{\resolve{\vhole{\thole}}{\vsu}}{\resolve{\thole}{\tsu}}$.
  %
  By Lemma~\ref{lem:vsu-ext} we know that
  $\tsu_{i} = \tsu_{i-1}[\thole_1 \mapsto t_1] \ldots [\thole_n \mapsto t_n]$.
  %
  We will assume, without loss of generality, that
  $\tsu_{i} = \tsu_{i-1}[\thole' \mapsto t']$.
  %
  Since $\tsu_{i-1}$ and $\tsu_{i}$ differ only in $\thole'$ but the resolved
  types differ,
  $\ptype{\trace_{i-1}}{f} = \intctx{\thole'}$
  and
  $\ptype{\trace_i}{f} = \intctx{t'}$.
  % $\resolve{\thole}{\tsu_{i-1}} = \intctx{\thole'}$
  % and
  % $\resolve{\thole}{\tsu_{i}} = \intctx{t'}$.
  %
  By Lemma~\ref{lem:context-compat},
  let $s$ be a concrete type such that $\intctx{s} = t$.
  %
  We show by case analysis on the evaluation rules that
  %
  $$\step{e_{i-1}}{\vsu_{i-1}}{\tsu_{i-1}[\thole' \mapsto s]}{\stuck}{\vsu}{\tsu}$$

  Finally, by Lemma~\ref{lem:resolve-compat} we know that
  % $\vsub{\resolve{\thole}{\tsu_{i-1}}}{\resolve{\ehole}{\vsu_{i-1}}}$,
  $\vsub{\ptype{\trace_{i-1}}{f}}{\resolve{\vhole{\thole}}{\vsu_{i-1}}}$
  and thus $\vsub{\intctx{\thole'}}{\resolve{\vhole{\thole}}{\vsu_{i-1}}}$.
  %
  Let
  $$
  \begin{array}{lcl}
  u &=& \gen{s}{\tsu} \\
  v &=& \resolve{\vhole{\thole}}{\vsu_{i-1}}\sub{\ehole'[\thole']}{u}\sub{\thole'}{s} \\
  \end{array}
  $$
  $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$ in $i$ steps.
  %and
  % $\vsub{\resolve{\thole}{\tsu_{i}}}{\resolve{\ehole}{\vsu_{i}}}$.
  % \vcompat{\resolve{\vhole{\thole}}{\vsu_{i-1}}}{\resolve{\thole}{\tsu_{i-1}}}
  % =
  % \tcompat{\ptype{i-1}{f}}{t}
  % \\
  % \vcompat{\resolve{\vhole{\thole}}{\vsu_{i}}}{\resolve{\thole}{\tsu_{i}}}
  % =
  % \tincompat{\ptype{i}{f}}{t}
  %
  % Finally, by the definition of $\typeofsym$,
  % we know that $\resolve{\vhole{\thole}}{\vsu_{i-1}} = \invctx{v}$
  % such that $\vsub{T'[\thole']}{v}$.
  %such that $\resolve{\typeof{v}}{\tsu_{i-1}} = T'[\thole']$
  %
  %\ES{need to connect \ptype{\trace{i}}{f} and \resolve{\vhole{\thole}}{\vsu_i}}
  %
  % Now, consider any value $u$ such that $\vcompat{\invctx{u}}{t}$,
  % we show by case analysis on the evaluation relation that
  % $\invctx{u}$ could not make progress at this step, and thus
  % $\steps{\eapp{f}{\invctx{u}}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
\end{proof}

@@

\begin{proof}[\textbf{Proof of Theorem~\ref{thm:soundness}}]
%
Suppose $\trace$ witnesses that $f$ gets stuck,
and let $s = \ptype{\trace}{f}$.
We show that \emph{all} types $t$ have stuck-inducing
values by splitting cases on whether $t$ is
compatible with $s$. %the partial type upto $\trace$.
%
\begin{description}
\item [Case \tcompat{s}{t}:]
  Let $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{\stuck}{\vsu}{\tsu}$.
  %
  The value $v = \resolve{\vhole{\thole}}{\vsu}$ demonstrates that
  $\eapp{f}{v}$ gets stuck.
\item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, we can derive
  a $v$ from $\trace$ such that $\hastype{v}{t}$ and $\eapp{f}{v}$ gets
  stuck.
% \item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, every $v$
%   such that \hastype{v}{t} demonstrates that $\eapp{f}{v}$ gets stuck.
%   % \ES{do we need to say anythign else?}
\end{description}
\end{proof}
